Rapid shortening at the eastern margin of the Tibetan plateau prior to the 2008 Mw=7.9 Wenchuan earthquake

### 2014-09-11 paper.tex
Test modifications.
Introduction edits.

\title{Rapid shortening at the eastern margin of the Tibetan plateau prior to the 2008 Mw=7.9 Wenchuan earthquake}
\author{T. Ben Thompson, Brendan J. Meade}
\date{September 2014}
\usepackage{graphicx}

@@

\section{Abstract}
The Longmen Shan is the steepest topographic front of the India-Asia collision and was the site of the Mw=7.9 Wenchuan earthquake.
Shortening estimates across the Longmen Shan provide strain accumulation rates and clarify the eastward extrusion of the Tibetan plateau.
Here, to explain the interseismic GPS velocities across the greater Longmen Shan region, we develop a boundary element model including earthquake cycle effects, topography, the westward dipping Beichuan fault, and a ~20 km deep, shallowly dipping, detachment.
The detachment is inferred from observations of slip during and after the Wenchuan earthquake and from structural considerations.
Previous analyses which neglected the detachment and earthquake cycle effects have found shortening rates near zero.
In contrast, we find that interseismic GPS data are consistent with a shortening rate of 6$\pm$2 mm/yr.
These results suggest that the Longmen Shan is an active fold-and-thrust belt with Wenchuan style earthquake recurrence intervals of <600 years.

\section{Introduction}
The Longmen Shan range is located on the eastern margin of the Tibetan and rises 4000m over 30km (CHECK THESE NUMBERS ON GOOGLE EARTH).
The 2008 $\mathrm{M_w}$ 7.9 Wenchuan earthquake ruptured 250km along strike on the Beichuan and Pengguan faults, resulting in approximately XXXXX fatalities (CITE STONE)
Prior geodetic estimates suggest almost zero horizontal shortening (Chen 2000, Shen 2005, etc.), in contrast with the dominant thrust slip-sense during the Wenchuan rupture. 
The lack of shortening is reconciled with the steep topography by appealing to a theory of lower crustal inflation whereby high topography can be formed without upper crustal shortening (CITE clark, royden, england, molnar, kirby, etc)
However, the magnitude and location of the Wenchuan earthquake suggest a more typical fold-and-thrust belt.
Using seismic reflection data, Hubbard and Shaw find almost 100\% shortening in the range-front 
Following on these works, there is a growing body of literature supporting the interpretation of the Longmen Shan as a fold-and-thrust belt (CITE, CITE, CITE) as opposed to a product of viscous lower crustal inflation.

Consistent with the fold-and-thrust mechanism
However, it remains to be explained how slow present-day geodetic shortening estimates can be reconciled with the Wenchuan earthquake and the large structurally-observed shortening.
Here, we present a model that explains the regional geodetic velocities as the effect of a complex interseismically locked fault geometry, including a 20km deep detachment as well as the range-front thrust Beichuan fault.
This geometry is consistent with the structural interpretations of Hubbard,Shaw,Klinger, with coseismic and postseismic Wenchuan slip distributions that attribute significant motion to a deep detachment (CITE QI, FIELDING).  
Also, seismic observations of P/sP conversion in the crust are suggestive of a low velocity layer near 20km depth (CITE ZHANG ET. AL. CRUSTAL STRUCTURE). 
Using this range-front thrust plus detachment geometry, we infer an average horizontal shortening.
Next, we show how this geometry can drives the steep velocity gradients far to the northwest of the range front, thus explaining the absence of a range-front interseismic velocity gradient.
Finally, we discuss the effects of Longmen Shan fault geometry on regional seismic hazard.

The Longmen Shan are a mountain range at boundary between the Tibetan Plateau and the Sichuan Basin. West of the Longmen Shan, in the eastern Tibetan plateau, surface GPS velocity vectors rotate from northeasterly to easterly or south-easterly. 
This rotation has been attributed to either the extrusion of crustal blocks or topographically driven viscous flow impacting a cold, elastically and viscously rigid, Sichuan basin.
Because of its uniquely steep topographic gradient, lack of foreland basin, etc..., the Longmen Shan is a natural testing ground for these hypotheses.
The steep topographic front of the Longmen Shan has been explained by both horizontal shortening (cite meade, hubbard/shaw) and lower crustal inflation (cite clark, royden, england, molnar, others). 
The contrast between high topography and low shortening rates is a central point in this debate.
Cite kirby
The Mw 7.9 2008 Wenchuan earthquake ruptured 250km NE-SW along the Pengguan and Beichuan faults in the Longmenshan (cite xu 2008);e .
Average slip was 4m 
80,000 fatalities. How much slip.
Using InSAR measurements and post-Wenchuan GPS observations, Qi et. al. 2011 found evidence of 2-6m of slip on a 20km deep detachment that extends 60km northwest underneath the Tibetan Plateau. 
Our goal is to demonstrate that locking on this deep detachment is observable in the interseismic GPS field prior to the Wenchuan earthquake.
Further, we will show how a deep detachment in the Longmen Shan shifts the main velocity gradient far to the northwest of the Beichuan fault trace.
Hence, the detachment obscures the near-field geodetic shortening signature. 
Accounting for the detachment, we present increased estimates of the shortening rate in the Longmen Shan.
Fielding and .. (CITE) interpret high resolution satellite gravity observations to indicate that the Tibetan Plateau is being thrust over the Sichuan Basin.

>>>>

\title{Rapid shortening at the eastern margin of the Tibetan plateau prior to the 2008 Mw=7.9 Wenchuan earthquake}
\author{T. Ben Thompson, Brendan J. Meade}
\date{September 2014}
\usepackage{graphicx}
\usepackage[margin=1.0in]{geometry}

@@

\section{Abstract}
The Longmen Shan is the steepest topographic front of the India-Asia collision and was the site of the Mw=7.9 Wenchuan earthquake.
Shortening estimates across the Longmen Shan provide strain accumulation rates and clarify the eastward extrusion of the Tibetan plateau.
Here, to explain the interseismic GPS velocities across the greater Longmen Shan region, we develop a boundary element model including earthquake cycle effects, topography, the westward dipping Beichuan fault, and a ~20 km deep, shallowly dipping, detachment.
The detachment is inferred from observations of slip during and after the Wenchuan earthquake and from structural considerations.
Previous analyses which neglected the detachment and earthquake cycle effects have found shortening rates near zero.
In contrast, we find that interseismic GPS data are consistent with a shortening rate of 6$\pm$2 mm/yr.
These results suggest that the Longmen Shan is an active fold-and-thrust belt with Wenchuan style earthquake recurrence intervals of $<$600 years.

\section{Introduction}
The Longmen Shan range is located on the eastern margin of the Tibetan and rises 4000m over 30km (CHECK THESE NUMBERS ON GOOGLE EARTH).
The 2008 $\mathrm{M_w}$ 7.9 Wenchuan earthquake ruptured 250km along strike on the Beichuan and Pengguan faults, resulting in approximately XXXXX fatalities (CITE STONE)
Prior geodetic estimates suggest almost zero horizontal shortening (Chen 2000, Shen 2005, Meade 2007, Loveless and Meade 2011), in contrast with the dominant thrust slip-sense during the Wenchuan rupture. 
Others authors reconcile the lack of shortening with the steep topography by appealing a lower crustal inflation tectonic model for the eastern Tibetan Plateau (CITE clark, royden, burchfiel, kirby, etc).
However, the magnitude and location of the Wenchuan earthquake suggest a more typical fold-and-thrust belt. 
Using seismic reflection data, Hubbard and Shaw 2009 find a well-developed fold-and-thrust fault system geometry with long-term shortenings up to 100\%. 
A growing body of literature supports this interpretation, 
Fielding and McKenzie 2012 interpret satellite gravity-field data to infer that mass of the Longmen Shan is supported by a flexed Sichuan Basin. 
This fold-and-thrust structure may extend deep into the plateau as evidenced by seismic tomography and coseismic slip inversions (CITE ZHANG, QI, FIELDING). 
CITE THE PAPERS EXAMINING THE LONGRIBA FAULT THAT PROPOSE A DETACHMENT

However, it remains to be explained how slow present-day geodetic shortening estimates can be reconciled with the Wenchuan earthquake and the large structurally-observed shortening.
Prior estimates can be categorized into those that neglect earthquake cycle effects, analyzing the fault slip-deficit without no explicit model of the geometry or tectonics, (CHEN, SHEN) and those that use block models, which normally assume a simplified fault geometry. 
Here, we present a model that includes earthquake cycle effects to explain the regional geodetic velocities as the result of a complex interseismically locked fault geometry, including a 20km deep detachment and the range-front thrust Beichuan fault.
This geometry is consistent with structural interpretations and with the geometry of Qi et. al 2012 and Fielding et. al 2013. 
Using this range-front thrust plus detachment geometry, we infer an average horizontal shortening between the TIbetan Plateau.
Next, we show how this geometry can drives the steep velocity gradients far to the northwest of the range front, thus explaining the absence of a range-front interseismic velocity gradient.
Finally, we discuss the effects of Longmen Shan fault geometry on regional seismic hazard and tectonics.

Maybe insert this information? --- Using InSAR measurements and post-Wenchuan GPS observations, Qi et. al. 2011 found evidence of 2-6m of slip on a 20km deep detachment that extends 60km northwest underneath the Tibetan Plateau.
